---
title: "Computer AidedDetection and DIagnosis of Breast Cancer"
author: "Sanja Stanisic, Universita' degli Studi di Milano Bicocca, CdLM Data Science, n. 800409"
date: "17. April 2021."
output:
  word_document:
    toc: yes
    toc_depth: '5'
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '5'
  ioslides_presentation:
    css:
    - css/fonts.css
    - css/custom.css
    - css/title-slide.css
    - css/slide-background.css
    includes:
      before_body: html/title.html
    toc: yes
    transition: default
    widescreen: yes
  beamer_presentation:
    colortheme: lily
    fig_caption: no
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    theme: Hannover
    toc: yes
  pdf_document:
    toc: yes
    toc_depth: 5
  slidy_presentation:
    highlight: default
  prettydoc::html_pretty:
    df_print: paged
    highlight: vignette
    theme: architect
    toc: yes
    toc_depth: 5
course: Data Science Lab
link-citations: true
bibliography: biblio.bib
---

```{r setup, include=FALSE}
# Use 'verbatim = TRUE' as chunk option to show chunk code as is
if(require(knitr)==FALSE) install.packages("knitr")
hook_source_def = knit_hooks$get('source')
knit_hooks$set(source = function(x, options){
  if (!is.null(options$verbatim) && options$verbatim){
    opts = gsub(",\\s*verbatim\\s*=\\s*TRUE\\s*", "", options$params.src)
    bef = sprintf('\n\n    ```{r %s}\n', opts, "\n")
    stringr::str_c(bef, paste(knitr:::indent_block(x, "    "), collapse = '\n'), "\n    ```\n")
  } else {
     hook_source_def(x, options)
  }
})
```

## Data for analyze 


```{r message=FALSE, warning=FALSE, include=TRUE}
library(data.table)
library(dplyr)

filePath <- "results-p03-full.csv"

all_dt <- fread(filePath, stringsAsFactors = TRUE, dec=".")

tibble(all_dt)

```

This dataset consists of 1333 instances with 15 features:


```{r message=FALSE, warning=FALSE, include=TRUE}

all_dt$PatientId <- NULL
all_dt$FullFilePath <- NULL
all_dt$ROIFilePath <- NULL

all_dt$LeftOrRightBrest <-  as.numeric(all_dt$LeftOrRightBrest)
all_dt$LeftOrRightBrest <- NULL

all_dt$ImageView <- as.numeric(all_dt$ImageView)
all_dt$ImageView <- NULL

all_dt$TestOrTraining <- as.numeric(all_dt$TestOrTraining)
all_dt$TestOrTraining <- NULL

all_dt$CalcType <- as.numeric(all_dt$CalcType)
all_dt$CalcDistribution <- as.numeric(all_dt$CalcDistribution)
all_dt$Patology <- as.numeric(all_dt$Patology)

head(all_dt)

```

### Preprocessing

Firstly, data should be set in adequate format.


```{r message=FALSE, warning=FALSE, include=TRUE}

all_dt_ex <- fread(filePath, stringsAsFactors = TRUE, dec=".")
all_dt_ex$PatientId <- NULL
all_dt_ex$FullFilePath <- NULL
all_dt_ex$ROIFilePath <- NULL

all_dt_ex$LeftOrRightBrest <-  as.numeric(all_dt_ex$LeftOrRightBrest)
all_dt_ex$LeftOrRightBrest <- cut(all_dt_ex$LeftOrRightBrest, 2, labels=c('LEFT', 'RIGHT'))
all_dt_ex$LeftOrRightBrest <- NULL

all_dt_ex$ImageView <- as.numeric(all_dt_ex$ImageView)
all_dt_ex$ImageView <- cut(all_dt_ex$ImageView, 2, labels=c('CC', 'MLO'))
all_dt_ex$ImageView <- NULL

all_dt_ex$TestOrTraining <- NULL

tibble(all_dt_ex)
```


After that, it should be checked is there missing values in dataset.

```{r echo=FALSE, message=TRUE, warning=FALSE, include=TRUE}
colSums(is.na(all_dt)) %>% show() # 0 in respective coumn indicates that there are no missing data
```
Obtained result indicate that thete is no misisng values. Therefore, there is no need to correct existng data.

### Data exploration

Since the research question is to predict if the patient has malignant changes, so variable "pathology" to be the dependent variable in this analysis. That variable is treated as a discrete attribute and its prediction will be executed as classification process.



Firstly, distribution of "pathology" is examinated.

```{r message=FALSE, warning=FALSE, echo=FALSE}

print("pathology")
all_dt$Patology %>% table()

library(ggplot2)

ggplot(all_dt,aes(x =Patology)) + 
  geom_bar(width = 0.4,fill =c("green","red")) + 
  geom_text(stat = 'count', aes(label =..count..),vjust =-0.5) + 
  theme_bw() + 
  theme_classic() + 
  ylab("number of people") + 
  ggtitle("Pathology distribution") 

```


Corelation among variables in dataset is calculated and displayed on the following diagram. 

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(data.table)
library(tidyverse)

all_df <- as.data.frame(all_dt) 

M <- cor(all_df)
M1 <- round(M, 3)
M1
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(corrplot)

M <- cor(all_df)
corrplot(M, method = "circle")
```

In this diagram, positive corelation is marked with different shades of blue, while negative correlation is marked with different shades of red. More intesive color indicate that correlation is higher.

```{r message=FALSE, warning=FALSE, echo=FALSE}

library(lares)

corr_var(all_df, # name of dataset
  Patology, # name of variable to focus on
  top = 5 # display top 5 correlations
)
```


### Models

Different Machine Learning models were chosen for predicting the "target" variable. Here is the list of models that are used in this report:

- k Nearest Neihbours (**k-nn**), described in [@murphy pp. 16--18]. An object is classified by a plurality vote of its neighbors, with the object being assigned to the class most common among its k nearest neighbors (k is a positive integer, typically small). I 

- Naive Bayes (**nb**), explained in [@murphy pp. 82--95]. It is simple "probabilistic classifier" based on applying Bayes' theorem, with strong (e.g. naïve) assumptions of independence between the features. In other words, naive Bayes classifier assume that the value of a particular feature is independent of the value of any other feature, given the class variable.

- SVM with Linear Kernel (**svm-l**), described in [@murphy pp.  482--486]. Training algorithm of SVM builds a model that assigns new examples to one category or the other, making it a non-probabilistic binary linear classifier.

- SVM with Radial Kernel (**svm-r**), also described in [@murphy pp. 498--505]. It is using the kernel trick, which implicitly maps kernel inputs into high-dimensional feature spaces where features are linearly separable. In this case kernel is defined with Gaussian radial basis function, given by formula: 

$$k(x_{i}, x_{j}) = e ^ {- \sigma {|x_{i} - x_{j}|}^2 }$$ 


- Random Forest (**rf**), also described in [@murphy pp. 550--553]. Random forest operate by constructing a multitude of decision trees at training time and outputting the value that is mean/average prediction of the individual trees.


### Implementation and evaluation

It is clear that various different alternatives and experiments should be created during ML process implementation.

Because of its popularity, efficiency, simplicity and flexibility and because of author's previous experience, R language and environment for statistical computing and graphics [@rfoundations] is used to implement the ML process. A decision tree is a flowchart-like structure in which each internal node represents a "test" on an attribute, each branch represents the outcome of the test, and each leaf node represents a class label (decision taken after computing all attributes). It is clear that  paths from root to leaf represent classification rules.

The following ML predictor models are developed with R functions:

- Function 'knn'[@rfunknn] in library 'class'[@rbibclass] is used for k-nn model realization.

- Function 'NaiveBayes'[@rfunnaivebayes] in library 'klaR'[@rbibklar] is used for nb model realization.

- Function 'ksvm' [@rfunkvsm] in library 'kernlab'[@rbibkernlab] with parameter kernel = vanilladot() that represents linear kernel, is used for svm-l model realization.

- Function 'ksvm'in library 'kernlab' with parameter kernel = "rbfdot" - which represents radial kernel, is used for svm-r model realization.

- Function 'randomForest'[@rfunrandomforest] in library 'randomForest'[@rbibrandomforest] is used for rf model realization.

Last, but not the least, R function 'train' [@rfuntrain] in library 'caret' [@rbibcaret] is used as umbrella that covers all the previously mentioned R functions and libraries for ML. They enables handling of a various learning models and functions in a uniform manner. In this moment, more than 230 classification and regression models are 'out-of-a-box' available for use with 'caret' and all of them are enlisted in [@kuhnam].

Developed models are compared using k-fold validation [@murphy pp. 201--210], with value of parameter k is set to 10. Selected 10-fold validation is realized with caret R functions. In order to achieve exactly the same conditions for comparison among developed ML methods, in all 10-fold validation scenarios, random generator is set on predefined value 155294099.

In order to evaluate quality of the selected ML regression methods, various measures [@murphy pp. 176--194] are used.

The following overall measures are calculated for ML models:

For measuring the performance of algorithms, sensitivity (or recall), specificity and accuracy were used because these three criteria are used more in the medical field.

For calculation of sensitivity, specificity and accuracy confusion matrix is required. In the following table, a confusion matrix is shown:

|                          | Actual class is C1    | Actual class is C2     |
| :----------------------- | :-------------------: | ---------------------: |
|  Predicted class is *C1* | True positive ($TP$)  | False positive ($FP$)  |
|  Predicted class is *C2* | False negative ($TN$) | True negative ($TN$)   |

Cells in confusion matrix have the following meaning [@alizedah2019]:
 - Actual class is the class which determined by angiography and it is existed in dataset. 
 - Predicted class is the one which is predicted by algorithms.
 - $TP$ is number of samples of class C1 which has been correctly classified.
 - $TN$ is number of samples of class C2 which has been correctly classified.
 - $FN$ is number of samples of class C1 which has been falsely classified as C2.
 - $FP$ is number of samples of class C2 which has been falsely classified as C1.

According to confusion matrix, sensitivity, specificity and accuracy are calculated as follows:


$$ Specificity = \frac{TN}{TN+FP} $$

$$ Sensitivity = \frac{TP}{TP+FN}  $$
$$ Accuracy = \frac{TP+TN}{TP+TN+FP+FN}  $$


Quality of the classification algorithm is often displayed by ROC (receiver operating characteristic) curve. It is a diagram  showing the performance of a classification model at all classification thresholds. This curve plots two parameters true positive rate ($TPR$) and false positive rate ($FPR$).

True Positive Rate ($TPR$) is a synonym for recall and is defined as follows:

$$ TPR = \frac{TP}{TP+FN}  $$

False Positive Rate ($FPR$) is defined as follows:

$$ FPR = \frac{FP}{TN+FP}  $$

An ROC curve plots $TPR$ vs. $FPR$ at different classification thresholds. Lowering the classification threshold classifies more items as positive, thus increasing both false positives and true positives. 

Area Under the ROC Curve (AUC) measures the entire two-dimensional area underneath the entire ROC curve (think integral calculus) from (0,0) to (1,1).

```{r echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Prepare data set for cross validation and ROC calculation
all_dt$Class <- factor( ifelse(all_dt$Patology==1, 'BENIGN', 'MALIGNANT') )
all_dt$Patology <- NULL
```

```{r echo=FALSE, message=FALSE, include=FALSE}

library(caret)

train_control <- trainControl(
  method="repeatedcv", 
  number=10,
  repeats = 10,
  classProbs = TRUE,  
  summaryFunction = twoClassSummary)

```


```{r echo=FALSE, message=FALSE, warning=FALSE}
# Train the k-nn model:

library(class)

set.seed(155294099)
model_k_nn <- train(
  Class~., 
  data=all_dt, 
  trControl=train_control,
  metric = "ROC",
  method="knn",
  na.action = na.omit
  )
```

Display info about k-nn model after 10-fold validation:
```{r echo=FALSE, message=FALSE, warning=FALSE}
#model_k_nn$method
model_k_nn$finalModel
model_k_nn$resample$ROC
confusionMatrix(model_k_nn)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cm <- confusionMatrix(model_k_nn)

ct <- cm$table 
ct1 <- round(ct, digits=2)
fourfoldplot(ct1, color = c("darkblue", "lightblue"), conf.level = 0, margin = 1, main = "Confusion Matrix k-nn")

TP <- ct[1,1]
FP <- ct[1,2]
FN <- ct[2,1]
TN <- ct[2,2]

Specificity <- TN/(TN+FP)

Sensitivity <- TP/(TP+FN)

Accuracy <- (TP+TN)/(TP+TN+FP+FN)

Specificity
Sensitivity
Accuracy
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Train the nb model:

library(klaR)

set.seed(155294099)
model_nb <- train(
  Class~., 
  data=all_dt, 
  trControl=train_control, 
  metric = "ROC",
  method="nb",
  na.action = na.omit
  )
```

Display info about nb model after 10-fold validation:
```{r echo=FALSE, message=FALSE, warning=FALSE}
#model_nb$method
model_nb$finalModel
model_nb$resample$ROC
confusionMatrix(model_nb)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
cm <- confusionMatrix(model_nb)

ct <- cm$table 
ct1 <- round(ct, digits=2)
fourfoldplot(ct1, color = c("darkblue", "lightblue"), conf.level = 0, margin = 1, main = "Confusion Matrix nb")

TP <- ct[1,1]
FP <- ct[1,2]
FN <- ct[2,1]
TN <- ct[2,2]

Specificity <- TN/(TN+FP)

Sensitivity <- TP/(TP+FN)

Accuracy <- (TP+TN)/(TP+TN+FP+FN)

Specificity
Sensitivity
Accuracy
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
# Train the svm-l model:

library(kernlab)

set.seed(155294099)
model_svm_l <- train(
  Class~., 
  data=all_dt, 
  trControl=train_control, 
  metric = "ROC",
  method="svmLinear",
  na.action = na.omit
  )
```

Display info about svm-l model after 10-fold validation:
```{r echo=FALSE, message=FALSE, warning=FALSE}
#model_svm_l$method
model_svm_l$finalModel
model_svm_l$resample$ROC
confusionMatrix(model_svm_l)
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
cm <- confusionMatrix(model_svm_l)

ct <- cm$table 
ct1 <- round(ct, digits=2)
fourfoldplot(ct1, c("darkblue", "lightblue"), conf.level = 0, margin = 1, main = "Confusion Matrix svm-l")

TP <- ct[1,1]
FP <- ct[1,2]
FN <- ct[2,1]
TN <- ct[2,2]

Specificity <- TN/(TN+FP)

Sensitivity <- TP/(TP+FN)

Accuracy <- (TP+TN)/(TP+TN+FP+FN)

Specificity
Sensitivity
Accuracy
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
# Train the svm-r model:

library(kernlab)

set.seed(155294099)
model_svm_r <- train(
  Class~., 
  data=all_dt, 
  trControl=train_control, 
  metric = "ROC",
  method="svmRadialCost",
  na.action = na.omit
  )

```

Display info about svm-r model after 10-fold validation:
```{r echo=FALSE, message=FALSE, warning=FALSE}
#model_svm_r$method
model_svm_r$finalModel
model_svm_r$resample$ROC
confusionMatrix(model_svm_r)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cm <- confusionMatrix(model_svm_r)

ct <- cm$table 
ct1 <- round(ct, digits=2)
fourfoldplot(ct1, color = c("darkblue", "lightblue"), conf.level = 0, margin = 1, main = "Confusion Matrix svm-r")

TP <- ct[1,1]
FP <- ct[1,2]
FN <- ct[2,1]
TN <- ct[2,2]

Specificity <- TN/(TN+FP)

Sensitivity <- TP/(TP+FN)

Accuracy <- (TP+TN)/(TP+TN+FP+FN)

Specificity
Sensitivity
Accuracy
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Train the rf model:

library(randomForest)

set.seed(155294099)
model_rf <- train(
  Class~., 
  data=all_dt, 
  trControl=train_control, 
  metric = "ROC",
  method="rf",
  na.action = na.omit
  )

```

Display info about rf model after 10-fold validation:
```{r echo=FALSE, message=FALSE, warning=FALSE}
#model_rf$method
model_rf$finalModel
model_rf$resample$ROC
confusionMatrix(model_rf)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
cm <- confusionMatrix(model_rf)

ct <- cm$table 
ct1 <- round(ct, digits=2)
fourfoldplot(ct1, color = c("darkblue", "lightblue"), conf.level = 0, margin = 1, main = "Confusion Matrix rf")

TP <- ct[1,1]
FP <- ct[1,2]
FN <- ct[2,1]
TN <- ct[2,2]

Specificity <- TN/(TN+FP)

Sensitivity <- TP/(TP+FN)

Accuracy <- (TP+TN)/(TP+TN+FP+FN)

Specificity
Sensitivity
Accuracy
```



Let us display ROC curves for the all created ML methods:
```{r echo=FALSE, message=FALSE, warning=FALSE}

library(pROC)

set.seed(2)
ind = sample(2, nrow(all_dt), replace = TRUE, prob=c(0.9,0.1))
trainset = as.data.frame( all_dt[ind == 1,] )
testset = as.data.frame( all_dt[ind == 2,] )


k_nn_prediction = predict(model_k_nn, 
                     testset[,! names(testset) %in% c("Class")], 
                     type = "prob") 
k_nn_ROC = roc(response = head(testset, nrow(k_nn_prediction))[,c("Class")],
               predictor = k_nn_prediction$MALIGNANT,
               levels = levels(testset[ ,"Class"]))
plot(k_nn_ROC, type="S", col="red")


nb_prediction = predict(model_nb, 
                     testset[,! names(testset) %in% c("Class")], 
                     type = "prob") 
nb_ROC = roc(response = head(testset, nrow(nb_prediction))[,c("Class")],
               predictor = nb_prediction$MALIGNANT,
               levels = levels(testset[ ,"Class"]))
plot(nb_ROC, add=TRUE, col="orange")


svm_l_prediction = predict(model_svm_l, 
                     testset[,! names(testset) %in% c("Class")], 
                     type = "prob") 
svm_l_ROC = roc(response = head(testset, nrow(svm_l_prediction))[,c("Class")],
               predictor = svm_l_prediction$MALIGNANT,
               levels = levels(testset[ ,"Class"]))
plot(svm_l_ROC, add=TRUE, col="blue")


svm_r_prediction = predict(model_svm_r, 
                     testset[,! names(testset) %in% c("Class")], 
                     type = "prob") 
svm_r_ROC = roc(response = head(testset, nrow(svm_r_prediction))[,c("Class")],
               predictor = svm_r_prediction$MALIGNANT,
               levels = levels(testset[ ,"Class"]))
plot(svm_r_ROC, add=TRUE, col="darkblue")

rf_prediction = predict(model_rf, 
                     testset[,! names(testset) %in% c("Class")], 
                     type = "prob") 
rf_ROC = roc(response = head(testset, nrow(rf_prediction))[,c("Class")],
               predictor = rf_prediction$MALIGNANT,
               levels = levels(testset[ ,"Class"]))
plot(rf_ROC, add=TRUE, col="green")
title("ROC curves")
legend("topleft", legend=c("k-nn", "nb", "svm-l", "svm-r", "rf"),
       col=c("red", "orange", "blue", "darkblue", "green"),  
       lty=1:5, cex=0.8)

```



Compare ROC values of all models:
```{r echo=FALSE, message=FALSE, warning=FALSE}
model_list <- list("k-nn" = model_k_nn, "nb" = model_nb, "svm-l"= model_svm_l, "svm-r"= model_svm_r, "rf"= model_rf)
res <- resamples( model_list )
res$values
summary(res)
```

Obtained results can be visualized:

```{r echo=FALSE, message=FALSE, warning=FALSE}
dotplot(res, metric = "ROC")

bwplot(res, layout = c(3, 1))
```

Moreover, calculated ROC distributions should be visualized:
```{r echo=FALSE, message=FALSE, warning=FALSE}
res_df <- as.data.frame(res$values)
res_df$"k-nn~Sens" <- NULL
res_df$"k-nn~Spec" <- NULL
res_df$"nb~Sens" <- NULL
res_df$"nb~Spec" <- NULL
res_df$"svm-l~Sens" <- NULL
res_df$"svm-l~Spec" <- NULL
res_df$"svm-r~Sens" <- NULL
res_df$"svm-r~Spec" <- NULL
res_df$"rf~Sens" <- NULL
res_df$"rf~Spec" <- NULL
res_dt <- as.data.table(res_df)
res_dt <- melt(res_dt, id.vars = c('Resample'), 
     measure.vars = c('k-nn~ROC', 'nb~ROC', 'svm-l~ROC', 'svm-r~ROC', 'rf~ROC') )
res_dt[variable == 'k-nn~ROC', variable := 'k-nn']
res_dt[variable == 'nb~ROC', variable := 'nb']
res_dt[variable == 'svm-l~ROC', variable := 'svm-l']
res_dt[variable == 'svm-r~ROC', variable := 'svm-r']
res_dt[variable == 'rf~ROC', variable := 'rf']
res_dt%>%setnames("variable", "method")
ggplot(data=res_dt, aes(x=value, group=method, fill=method)) +
    geom_density(adjust=1.5, alpha=.4) +
    xlab("ROC") +
    ylab("value") + 
    scale_fill_manual(values=c("red", "orange", "blue", "darkblue", 'green')) +
    ggtitle("ROC distribution by model") 
```




### References




